---
title: "Chart code"
---

Here, we document some of the data and scripts used in producing the charts in the data visualization segments of the Institute. 

```{r}
# silently load packages
suppressPackageStartupMessages(library("data.table"))
suppressPackageStartupMessages(library("ggplot2"))

# set data.table printout options
options(
  datatable.print.nrows = 20,
  datatable.print.topn = 5,
  datatable.print.class = TRUE
)
```


## Correlation

Engineering students at 14 institutions persisting to year 4 and graduating by year 6, 1987--2019. 

Data are available in the [Institute repository data directory]( https://github.com/MIDFIELDR/2023-midfield-institute/tree/46d97d20593b59b7f2350689b41f2607eb8d0d91/data). If you download that CSV file and save it in your project data directory, you can input the data using:

```{r}
# import data
DT <- fread("data/four-year-six-year.csv")
DT
```

```{r}
ggplot(DT, aes(x = y4/1000, y = y6/1000)) +
  geom_smooth(method = "lm", se = FALSE, size = 0.6, color = "gray") +
  geom_point(size = 3, 
             shape = 21, 
             color = "black", 
             fill = "gray70") + 
  labs(x = "4-yr persisters (thousands)", 
       y = "6-yr completers (thousands)", 
       title = "")
```

```{r}
ggplot(DT, aes(x = y4/1000, y = y6/1000, color = sex, fill = sex)) +
  geom_smooth(method = "lm", se = FALSE, size = 0.6) +
  geom_point(size = 3, shape = 21) + 
  scale_color_manual(values = c("#1B7837", "#762A83")) +
  scale_fill_manual(values = c("#5AAE61", "#9970AB")) +
  labs(x = "4-yr persisters (thousands)", 
       y = "6-yr completers (thousands)", 
       title = "") + 
  theme(legend.position = "none") +
  geom_text(data = DT[institution == "B"],
            aes(x = y4/1000, y = y6/1000, label = sex),
            vjust = -1, hjust = 1, size = 4)
```

```{r}
ggplot(DT, aes(x = y4/1000, y = y6/1000, color = sex, fill = sex)) +
  geom_smooth(method = "lm", se = FALSE, size = 0.6) +
  geom_point(size = 3, shape = 21) + 
  scale_color_manual(values = c("#1B7837", "#762A83")) +
  scale_fill_manual(values = c("#5AAE61", "#9970AB")) +
  labs(x = "4-yr persisters (thousands)", 
       y = "6-yr completers (thousands)", 
       title = "") + 
  theme(legend.position = "none") +
  
  facet_wrap(vars(sex))
```

```{r}
dt <- copy(DT)
p <- ggplot(dt, aes(x = y4/1000, y = y6/1000, color = sex, fill = sex)) +
  superpose_point +
  superpose_color +
  superpose_fill_mid +
  # coord_equal() +
  facet_wrap(vars(reorder(institution, y6, median)),
             ncol = 7,
             as.table = FALSE)+
  labs(x = "4-yr persisters (thousands)",
       y = "6-yr completers (thousands)") +
  theme_graphclass(font_size = 20) +
  theme(legend.position = "none") +
  geom_text(data = dt[institution == "Institution G"],
            aes(x = y4/1000, y = y6/1000, label = sex),
            vjust = 0.5, hjust = c(-0.2, -0.4), size = 7) +
  scale_x_continuous(limits = c(-1, 19), breaks = seq(0, 15, 5)) +
  scale_y_continuous(limits = c(-1, 17), breaks = seq(0, 15, 5))
p
```

